Modelling multi-scale state-switching functional data with hidden Markov models

### 2021-09-16 Report/Sections/models.tex
Update Final Version

%Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$. 
The joint likelihood of both the parameters and the hidden states is given by the following:

@@

\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability. This formulation simplifies likelihood maximization by removing constraints in the optimization problem. We assume there are no covariate effects, but \cite{DeSouza:2017} and \citet{Adam:2019} both incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients. For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$. Figure \ref{fig:models} shows the dependence structure of an HMM.

@@

A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$ each taking integer values between $1$ and $N$. A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $\{X_1,\ldots, X_T\}$ and $\{Y_1,\ldots, Y_{t-1}\}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$. We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. 
%This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$. 
As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM. For example, since $f^{*(i,i^*)}$ depends on both $X_t = i$ and $X^*_{t,t^*} = i^*$, the distribution of $Y^*_{t,t^*}$ is defined by the pair $(X_t,X^*_{t,t^*})$. However, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$ and share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$. As a result, the distribution of the fine-scale observation $Y^*_{t,t^*}$ does not depend on $X_t$, but instead is defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency). %Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, not their fine-scale emission parameters $\theta^{*(\cdot,i)}$.

Due to the nested structure of the HHMM, %its likelihood is easily calculated using the forward algorithm.
it is straightforward to extend the forward algorithm and sum over all possible coarse-scale hidden states $X$ and all possible fine-scale hidden states $X^*$ in the likelihood.
%
Let $y$ be the sequence of observed coarse-scale emissions and
$y^* \equiv \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then

@@

After performing this transformation, the entire model must be redefined since the fine-scale HMM (or CarHMM) directly models the distribution of the summary statistics $\Z_t$ rather than the discretized curve $Y^*_t$. Because $\Z_t$ exists on a coarser scale than $Y^*_t$, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$. We denote these unobserved states as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g., conditional independence, auto-regressive structure between observations, etc.) Fine-scale model selection and validation should therefore be adjusted accordingly. %instead of $X^*_t$ and $Y^*_t$.

@@

Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
However, the fine-scale observations of a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods. As such, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
%
%Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here. 
For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, then this approach is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
The fine-scale process can also be modelled similarly to \citet{Langrock:2018}, who use B-splines to model the emission distribution of an HMM. This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}). 
Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. 
A continuous-time HMM may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. \citet{Xu:2018} model high-frequency biologging accelerometer data of individuals by incorporating a CTHMM into a hierarchical model similar to ours. However, they assume that individuals are partitioned into subgroups a priori whereas we use an HMM to infer the coarse-scale hidden states.

>>>>

The joint likelihood of both the parameters and the hidden states is given by the following:

@@

\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability. This formulation simplifies likelihood maximization by removing constraints in the optimization problem. We assume there are no covariate effects, but similarly to the work of \citet{DeSouza:2017} for independent states and \citet{Adam:2019} for Markovian states, one could incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients. For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$. Figure \ref{fig:models} shows the dependence structure of an HMM.

@@

A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$ each taking integer values between $1$ and $N$. A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $\{X_1,\ldots, X_T\}$ and $\{Y_1,\ldots, Y_{t-1}\}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$. We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. 
%
As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM. For example, since $f^{*(i,i^*)}$ depends on both $X_t = i$ and $X^*_{t,t^*} = i^*$, the distribution of $Y^*_{t,t^*}$ is defined by the pair $(X_t,X^*_{t,t^*})$. However, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$ and share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$. As a result, the distribution of the fine-scale observation $Y^*_{t,t^*}$ does not depend on $X_t$, but instead is defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).

Due to the nested structure of the HHMM, it is straightforward to extend the forward algorithm and sum over all possible coarse-scale hidden states $X$ and all possible fine-scale hidden states $X^*$ in the likelihood.
%
Let $y$ be the sequence of observed coarse-scale emissions and
$y^* \equiv \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then

@@

After performing this transformation, the entire model must be redefined since the fine-scale HMM (or CarHMM) directly models the distribution of the summary statistics $\Z_t$ rather than the discretized curve $Y^*_t$. Because $\Z_t$ exists on a coarser scale than $Y^*_t$, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$. We denote these unobserved states as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g., conditional independence, auto-regressive structure between observations, etc.) Fine-scale model selection and validation should therefore be adjusted accordingly.

@@

Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
However, the fine-scale observations of a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods. As such, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
%
For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, then this approach is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
The fine-scale process can also be modelled similarly to \citet{Langrock:2018}, who use B-splines to model the emission distribution of an HMM. This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}). 
Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. 
A continuous-time HMM may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. \citet{Xu:2018} model high-frequency biologging accelerometer data of individuals by incorporating a CTHMM into a hierarchical model similar to ours. However, they assume that individuals are partitioned into subgroups a priori whereas we use an HMM to infer the coarse-scale hidden states.

### 2021-09-16 Report/Sections/results.tex
Update Final Version

%To illustrate an application of this method and compare the candidate models, 
We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the SciPy package in Python \citep{Virtanen:2019}. Each model is fit 100 times using random initializations, and
%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation. If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values. If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values. 
we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}. Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory. Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses. Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and %researchers 
unpublished data collected by the authors suggest that %significant numbers of 
prey-capture events can occur near the surface. %\citep{Fortune:2020}. 
Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:model_checking}. Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the %\textit{forward-backward algorithm} \citep{Zucchini:2016}. 
forward-backward algorithm. This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations are contained in Section 6 and 7 of of supplement A. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, histograms corresponding to subdive state 3 show that $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution. This indicates the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
any given behaviour as subdive state 3
%the behaviour at the beginning and end of dives as subdive state 3 
(see Figures 11 and 12 of supplement A). In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are %approximately $50$ to $100$ percent 
significantly larger than those of the CarHHMM-DFT. The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A). This suggests %that including auto-correlation in the model significantly affects parameter estimates, %Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A). These findings suggest 
that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.

>>>>

We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the SciPy package in Python \citep{Virtanen:2019}. Each model is fit 100 times using random initializations, and we keep the parameter estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}. Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory. Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses. Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.

@@

The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives. No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters. However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and unpublished data collected by the authors suggest that prey-capture events can occur near the surface.
Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.

@@

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:model_checking}. Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the forward-backward algorithm. This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2. Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT. Analogous histograms corresponding to the fine-scale observations are contained in Section 6 and 7 of of supplement A. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, histograms corresponding to subdive state 3 show that $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution. This indicates the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
any given behaviour as subdive state 3 (see Figures 11 and 12 of supplement A). In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are significantly larger than those of the CarHHMM-DFT. The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A). This suggests that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.

### 2021-09-16 Report/main.tex
small changes
End of day Jan 6 2021
Updates from Overleaf
Update Final Version

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\input{header}
\input{defs}

\begin{document}
\firstpage{1}
\lastpage{25}
\jvol{xx}
\issue{yy}
\jyear{2020}
\jid{CJS}
\aid{???}
% The running head contains the author names
\rhauthor{BlindedA and BlindedB}
\copyrightline{Statistical Society of Canada}
\Frenchcopyrightline{Soci\'et\'e statistique du Canada}
% History: received and accepted dates
\received{\rec{9}{July}{2009}}
\accepted{\acc{8}{July}{2010}}

@@

% Title, authors, affiliations
%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\title[]{Modelling multi-scale state-switching functional data with hidden Markov models}
%\author{Evan Sidrow\authorref{1}}
%\author{Nancy Heckman\authorref{1}}
%\author{Sarah Fortune\authorref{2}}
%\author{Andrew Trites\authorref{3,4}}
%\author{Ian Murphy\authorref{1}}
%\author{Marie Auger-M\'eth\'e\authorref{1,4}}
%\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada}
%\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada}
%\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada}
%\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep state-switching\sep statistical ecology\sep time series.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
%Insert your abstract here; it should
%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
%are strongly discouraged, and citations should be avoided. 
%The title and the abstract should be concise and descriptive;
%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

\iffalse
Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs. We detail a general hierarchical approach based on HMMs which incorporates a wider class of models at fine scales. As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
\fi

@@

% Email address for corresponding author
%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
\section{DISCUSSION}
\input Sections/discussion.tex
%
\begin{ack}{ACKNOWLEDGEMENTS}
This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC). 
Marie Auger-M\'eth\'e thanks the NSERC Discovery and the Canadian Research Chair programs.
Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
\end{ack}
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
%\input Sections/appendix.tex
\input Sections/appendixnh.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory

>>>>

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
%\usepackage{lineno}
\input{header}
\input{defs}

%\linenumbers

\begin{document}
\firstpage{1}
\lastpage{34}
\jvol{xx}
\issue{yy}
\jyear{2021}
\jid{CJS}
\aid{???}
% The running head contains the author names
\rhauthor{Sidrow et al.}
\copyrightline{Statistical Society of Canada}
\Frenchcopyrightline{Soci\'et\'e statistique du Canada}
% History: received and accepted dates
\received{\rec{14}{January}{2021}}
\accepted{\acc{14}{September}{2021}}

@@

% Title, authors, affiliations
%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\title[]{Modelling multi-scale state-switching functional data with hidden Markov models}
\author{Evan Sidrow\authorref{*1}}
\author{Nancy Heckman\authorref{1}}
\author{Sarah M. E. Fortune\authorref{2}}
\author{Andrew W. Trites\authorref{3,4}}
\author{Ian Murphy\authorref{5}}
\author{Marie Auger-M\'eth\'e\authorref{1,4}}
\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada}
\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada}
\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada}
\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
\affiliation[5]{Department of Biostatistics, University of Florida, Gainesville, USA}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep state-switching\sep statistical ecology\sep time series.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}

@@

% Email address for corresponding author
\correspondingauthor[*]{\\\email{evan.sidrow@stat.ubc.ca}} % only put this in after acceptance
%
\section{INTRODUCTION}
\input Sections/intro.tex
%
\section{MODELS AND PARAMETER ESTIMATION}
\label{sec:models}
\input Sections/models.tex
%
\section{KILLER WHALE CASE STUDY}
\label{sec:data}
\input Sections/results.tex
%
\section{SIMULATION STUDY}
\label{sec:sim_study}
\input Sections/simulation.tex
%
\section{DISCUSSION}
\input Sections/discussion.tex
%
\begin{ack}{ACKNOWLEDGEMENTS}
All killer whale data was collected under University of British Columbia Animal Care Permit no. A19-0053 and Fisheries and Oceans Canada Marine Mammal Scientific License for Whale Research no. XMMS 6 2019.
This project was supported in part by a financial contribution from Fisheries and Oceans Canada and the Natural Sciences and Engineering Research Council of Canada (Whale Science for Tomorrow).
This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) as well as the support of Fisheries and Oceans Canada (DFO). 
Marie Auger-M\'eth\'e and Nancy Heckman thank the NSERC Discovery program, and Marie Auger-M\'eth\'e additionally thanks the Canadian Research Chair program.
Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
We are grateful to Dr. Joe Watson for his constructive suggestions.
\end{ack}
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{FIGURES AND TABLES}
\input Sections/figures.tex
%
%\CJShistory
